Water, High-Altitude Condensates, and Methane Depletion in the Atmosphere of the Warm Neptune WASP-107\lowercase{b}

### 2017-08-12 ms.tex
captions

@@

\begin{figure*}
\includegraphics[width = \textwidth]{Figures/lc.pdf}
\caption{FIXME}
\label{fig:lc}
\end{figure*}

@@

For the best fit white light curve, the root-mean-square (rms) residuals were 93 ppm (excluding the star spot crossing), which is somewhat larger than the expected shot noise of 50 ppm. We attribute the excess noise to loss of flux off the edge of the detector, which can occur if there is variation in the position or length of the spatial scan. There is no evidence for correlated noise in the residuals, so to account for the excess noise we simply increased the per-point uncertainties by a factor of 1.7 to achieve a reduced $\chi^2$ value of unity.  We then used the Markov chain Monte Carlo (MCMC) algorithm to estimate parameter uncertainties \citep{foremanmackey13}.  The chain had 50 walkers which each ran for $10^4$ steps with the first 10\% discarded as burn-in. We tested for convergence by dividing the chain in two halves and confirming that they gave consistent results. The transit time was $t_0 = 2457910.45407\pm6\mathrm{e}{-5}$ BJD$_\mathrm{TDB}$ and the planet/star radius was $r_p/r_s = 0.14399\pm0.00017$.

@@

The best fit light curves have a median reduced $\chi^2$ value of 1.16.  To ensure that we did not underestimate the uncertainty on the transit depths, we rescaled the photometric uncertainties for all spectroscopic channels such that the $\chi^2$ values are unity. We performed an MCMC fit to the light curves with \texttt{emcee}.  For each light curve we ran a fit with 50 walkers and 1000 steps per walker, and tested for convergence as we did for the white light curve. The median transit depths and $1\,\sigma$ uncertainties are given in Table\,\ref{tab:transit_depths}.

@@

\subsubsection{Transit Depths Redder than 1.62 $\mu$m}
The red edge of the transmission spectrum is of interest for our analysis because methane is expected to be the dominant absorber over water at wavelengths greater than $1.6\,\mu$m. Unfortunately, we find that the our reddest spectroscopic light curves ($>1.62\mu$m) are too poor in quality to robustly measure the transit depth. The residuals exhibit correlated noise and the fits have reduced $\chi^2_\nu>2$. The transit depths are also sensitive to which method we use to fit for instrument systematics.  We therefore opt not to report transit depths redder than the 1.62 $\mu$m channel.

@@

\begin{figure*}
\includegraphics[width = \textwidth]{Figures/spectrum.pdf}
\caption{FIXME}
\label{fig:spectrum}
\end{figure*}

@@

\subsection{Atmospheric Metallicity from Interior Structure Modeling}
Given WASP-107b's unusually low density, we sought to estimate the range of atmospheric metallicities that are consistent with the observed mass and radius.  We estimated the bulk metallicity of WASP-107b using the structure evolution modeling of \cite{thorngren16}.  These models assume a thermally inert heavy-element core with a convective envelope of additively mixed H/He \citep{saumon95} and heavy-element impurities.  The heavy elements were a 50-50 rock-ice mix from ANEOS \citep{thompson90}.  We evolved the planets in time using the atmospheric models of \cite{fortney07}.  The results are sensitive to assumptions about the stellar age, which is uncertain \citep[either $0.6\pm0.2$ to $8.3\pm4.3$ Gyr depending on model assumptions;][]{mocnik17}. We therefore ran two models, with uniform age priors of either $0.2-1.0$ or $1.0-13.8$ Gyr.  We used the published mass and radius estimates \citep[$0.12\pm0.01\,M_\mathrm{J}$, $0.94\pm0.02$;][]{anderson17}.  A more precise mass estimate is in preparation (B. Benneke, priv.\,comm.); however, we find that our results are not signficantly affected when we vary the assumed mass by $\pm3\,\sigma$ from the published value because the dominant uncertainties are the stellar age and core mass.

Based on these assumptions, we fit for envelope metallicity and core mass using an MCMC with uniform priors on both parameters.  The MCMC burned in for $10^3$ steps and then collected $4\times10^6$ samples, which we thinned to $10^5$.  The resulting envelope metal mass fractions were converted to metallicities by assuming the mean molecular weight of the metals was about 18 (the value for water), using the approach of \cite{fortney13}. Figure\,\ref{fig:metal_prior} shows the results.  We find that the inferred atmospheric metallicity and core mass are inversely proportional, as expected (the more massive the core, the lighter the envelope must be to maintain the observed radius). Larger core masses and envelope metallicities are allowed for the younger stellar age (when planets age, they cool and contract).  For the younger (older) stellar age, we find an upper limit on the core mass of FIXME and an upper limit on atmospheric metallicity of FIXME.

\begin{figure}
\includegraphics[width = 0.5\textwidth]{Figures/metalByMass.png}
\caption{FIXME}
\label{fig:metal_prior}
\end{figure}

\subsection{Retrieval}
\label{sec:retrieval}
We used the CHIMERA chemically-consistent retrieval tool described in \cite{kreidberg15b} to determine the planet's basic atmospheric properties from the transmission spectrum.   Briefly, CHIMERA solves the transmission geometry problem using the equations in \cite{brown01, tinetti12}.  We parameterize atmospheric composition with metallicity and carbon-to-oxygen ratio (C/O) under the assumption of thermochemical equilibrium using the NASA CEA routine \citep{gordon96} to compute the molecular abundances for H$_2$, He, H$_2$O, CH$_4$, CO, CO$_2$, NH$_3$, H$_2$S, Na, K, HCN, C$_2$H$_2$, TiO, VO, and FeH.    We recently upgraded the transmission model to use correlated-K opacities \citep{lacis91, molliere15, amundsen16} for the aforementioned gases generated from the pre-tabulated line-by-line cross section database described in \cite{freedman14}. The transmission forward model is coupled with the powerful PyMultiNest tool \citep{buchner16} to solve the parameter estimation and model selection problems.

>>>>

\author{Daniel Thorngren}
\affiliation{University of California Santa Cruz}
\nocollaboration

@@

\begin{figure*}
\includegraphics[width = \textwidth]{Figures/lc.pdf}
\caption{Broadband and spectrophotometric transit light curves for WASP-107b measured with \textit{HST}/WFC3 compared to best fit models (left) and residuals from the best fit (right). Annotations indicate the central wavelength and root-mean-square (rms) residuals for each light curve. A possible star-spot crossing feature is shaded in gray in the upper right; our systematic error correction removes this feature from the spectroscopic light curves.} 
\label{fig:lc}
\end{figure*}

@@

For the best fit white light curve, the root-mean-square (rms) residuals were 93 ppm (excluding the star spot crossing), which is somewhat larger than the expected shot noise of 50 ppm. We attribute the excess noise to loss of flux off the edge of the detector, which can occur if there is variation in the position or length of the spatial scan. There is no evidence for correlated noise in the residuals, so to account for the excess noise we simply increased the per-point uncertainties by a factor of 1.7 to achieve a $\chi^2_\nu$ value of unity.  We then used the Markov chain Monte Carlo (MCMC) algorithm to estimate parameter uncertainties \citep{foremanmackey13}.  The chain had 50 walkers which each ran for $10^4$ steps with the first 10\% discarded as burn-in. We tested for convergence by dividing the chain in two halves and confirming that they gave consistent results. The transit time was $t_0 = 2457910.45407\pm6\mathrm{e}{-5}$ BJD$_\mathrm{TDB}$ and the planet/star radius was $r_p/r_s = 0.14399\pm0.00017$.

@@

The best fit light curves have a median $\chi^2_\nu$ value of 1.16.  To ensure that we did not underestimate the uncertainty on the transit depths, we rescaled the photometric uncertainties for all spectroscopic channels such that the $\chi^2_\nu$ values are unity. We performed an MCMC fit to the light curves with \texttt{emcee}.  For each light curve we ran a fit with 50 walkers and 1000 steps per walker, and tested for convergence as we did for the white light curve. The median transit depths and $1\,\sigma$ uncertainties are given in Table\,\ref{tab:transit_depths}.

@@

\subsubsection{Transit Depths Redder than 1.62 $\mu$m}
The red edge of the transmission spectrum is of interest for our analysis because methane is expected to be the dominant absorber over water at wavelengths greater than $1.6\,\mu$m. Unfortunately, we find that the our reddest spectroscopic light curves ($>1.62\mu$m) are too poor in quality to robustly measure the transit depth. The residuals exhibit correlated noise and the fits have $\chi^2_\nu>2$. The transit depths are also sensitive to which method we use to fit for instrument systematics.  We therefore opt not to report transit depths redder than the 1.62 $\mu$m channel.

@@

\begin{figure*}
\includegraphics[width = \textwidth]{Figures/spectrum.pdf}
\caption{The transmission spectrum of WASP-107b (points with 1\,$\sigma$ error bars) compared to retrieved models (blue line with shaded $1$ and $2\sigma$ confidence intervals). The features at $1.15$ and $1.4$ $\mu$m are due to water absorption. The right-hand axis indicates the normalized atmospheric scale height assuming a $100\times$ solar composition.}
\label{fig:spectrum}
\end{figure*}

@@

\subsection{Atmospheric Metallicity from Interior Structure Modeling}
Given WASP-107b's unusually low density, we sought to estimate the range of atmospheric metallicities that are consistent with the observed mass and radius.  We estimated the bulk metallicity of WASP-107b using the structure evolution modeling of \cite{thorngren16}.  These models assume a thermally inert heavy-element core with a convective envelope of additively mixed H/He \citep{saumon95} and heavy-element impurities.  The heavy elements were a 50-50 rock-ice mix from ANEOS \citep{thompson90}.  We evolved the planets in time using the atmospheric models of \cite{fortney07}.  The results are sensitive to assumptions about the stellar age, which is uncertain \citep[either $0.6\pm0.2$ to $8.3\pm4.3$ Gyr depending on model assumptions;][]{mocnik17}. We therefore ran two models, with uniform age priors of either $0.2-1.0$ or $1.0-13.8$ Gyr.  We used the published mass and radius estimates \citep[$0.12\pm0.01\,M_\mathrm{J}$, $0.94\pm0.02$;][]{anderson17}.  A more precise mass estimate is in preparation (B. Benneke, priv.\,comm.); however, we find that our results are not significantly affected when we vary the assumed mass by $\pm3\,\sigma$ from the published value because the dominant uncertainties are the stellar age and core mass.

Based on these assumptions, we fit for envelope metallicity and core mass using an MCMC with uniform priors on both parameters.  The MCMC burned in for $10^3$ steps and then collected $4\times10^6$ samples, which we thinned to $10^5$.  The resulting envelope metal mass fractions were converted to metallicities by assuming the mean molecular weight of the metals was about 18 (the value for water), using the approach of \cite{fortney13}. Figure\,\ref{fig:metal_prior} shows the results.  We find that the inferred atmospheric metallicity and core mass are inversely proportional, as expected (the more massive the core, the lighter the envelope must be to maintain the observed radius). Larger core masses and envelope metallicities are allowed for the younger stellar age since planets cool and contract as they age.  For the younger (older) stellar age, we find an upper limit on the core mass of FIXME and an upper limit on atmospheric metallicity of FIXME.

\begin{figure}
\includegraphics[width = 0.5\textwidth]{Figures/metalByMass.png}
\caption{Estimated envelope metallicity and core mass based on interior structure modeling of the mass and radius of WASP-107b. FIXME ages.}
\label{fig:metal_prior}
\end{figure}

\subsection{Retrieval}
\label{sec:retrieval}
We also inferred the composition of the atmosphere directly from the transmission spectrum using the CHIMERA chemically-consistent retrieval tool described in \cite{kreidberg15b}.   Briefly, CHIMERA solves the transmission geometry problem using the equations in \cite{brown01, tinetti12}.  We parameterize atmospheric composition with metallicity and carbon-to-oxygen ratio (C/O) under the assumption of thermochemical equilibrium using the NASA CEA routine \citep{gordon96} to compute the molecular abundances for H$_2$, He, H$_2$O, CH$_4$, CO, CO$_2$, NH$_3$, H$_2$S, Na, K, HCN, C$_2$H$_2$, TiO, VO, and FeH.    We updated the transmission model to use correlated-K opacities \citep{lacis91, molliere15, amundsen16} for the aforementioned gases generated from the pre-tabulated line-by-line cross section database described in \cite{freedman14}. The transmission forward model is coupled with the powerful PyMultiNest tool \citep{buchner16} to solve the parameter estimation and model selection problems.
